---
title: "re_airquality"
author: "Emma Biland"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)

df <- datasets::airquality
```

## 1. Dataset description

The `airquality` dataset provides daily air quality measurements in New York City from May to September 1973. The main variable of interest is **Ozone** (ppb). Other variables include **Solar.R** (solar radiation in Langley), **Wind** (mph), **Temp** (Fahrenheit), and the **Month** and **Day** of the observation. Some values are missing (e.g. Ozone and Solar.R).

```{r preview}
knitr::kable(utils::head(df, 6), caption = "First six rows of the airquality dataset")
```

## 2. Research question

> **How does temperature influence ozone concentration, and are there seasonal (monthly) patterns in ozone levels?**

## 3. Statistical metrics

We begin by computing a few summary statistics to better understand the key variables.

```{r summary-metrics}
df_clean <- dplyr::filter(df, !is.na(Ozone), !is.na(Temp))

summary_stats <- dplyr::summarise(
  df_clean,
  mean_ozone = round(mean(Ozone), 1),
  max_ozone = max(Ozone),
  min_ozone = min(Ozone),
  mean_temp = round(mean(Temp), 1),
  n = dplyr::n()
)

knitr::kable(summary_stats, caption = "Basic statistics on Ozone and Temperature")
```

## 4. Figures and patterns

### a) Ozone concentration by month

```{r boxplot-month}
ggplot2::ggplot(df, ggplot2::aes(x = factor(Month), y = Ozone)) +
  ggplot2::geom_boxplot(fill = "#FFCC80", color = "#FF6F00", na.rm = TRUE) +
  ggplot2::labs(title = "Monthly distribution of ozone concentration",
                x = "Month (May to September)", y = "Ozone (ppb)") +
  ggplot2::theme_minimal()
```

### b) Temperature vs Ozone (scatterplot)

```{r scatter-temp-ozone}
ggplot2::ggplot(df_clean, ggplot2::aes(x = Temp, y = Ozone)) +
  ggplot2::geom_point(color = "#1565C0", alpha = 0.7) +
  ggplot2::geom_smooth(method = "lm", se = FALSE, color = "darkred") +
  ggplot2::labs(title = "Relationship between temperature and ozone",
                x = "Temperature (°F)", y = "Ozone (ppb)") +
  ggplot2::theme_light()
```

### c) Temporal trend in Ozone (by date)

```{r ozone-time-series}
df_date <- df %>%
  dplyr::filter(!is.na(Ozone)) %>%
  dplyr::mutate(date = base::as.Date(paste(1973, Month, Day, sep = "-")))

ggplot2::ggplot(df_date, ggplot2::aes(x = date, y = Ozone)) +
  ggplot2::geom_line(color = "#43A047") +
  ggplot2::labs(title = "Ozone concentration over time",
                x = "Date", y = "Ozone (ppb)") +
  ggplot2::theme_minimal()
```

## 5. Interpretation and conclusion

We observed that ozone concentration varies strongly between months, with the highest levels in July and August. A clear positive correlation is visible between temperature and ozone, suggesting that warmer days tend to have higher ozone levels — a pattern consistent with known photochemical processes.

Over time, ozone values show a noisy but slightly increasing trend toward midsummer. These results support the idea that temperature is a strong predictor of ozone, and that seasonality plays a significant role in air quality patterns.

Further exploration could include multivariate models controlling for solar radiation and wind speed to better isolate temperature effects.
